Legacy Data Confounds Genomics Studies

### 2018-10-25 legacy.tex
updated figures

Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques. 
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. For other analyses, removing suspicious SNPs may be adequate. \sgcomment{Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.}

@@

\section{Discussion}
Two of these studies, which identified SNPs associated to ADHD symptoms and metabolic traits respectively, used HapMap as a reference panel, rather than 1000 Genomes \citep{Ebejer2013}. \todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
%While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
%The data used was imputed genotype data using Hapmap II as the reference.
%One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ .
%While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
%Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
%We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.

@@

For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis. The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in \todo{number} of SNPs and  \todo{2,952,684} insertions and deletions that passed the same quality controls. We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.

\begin{table}[h]
\begin{tabular}{l  r r}
                      & {Repeat}  & {Non-Repeat}       \\ \hline
{SNP}  & \todo{number} & \todo{number}  \\  
{INDEL} & \todo{number}   & \todo{number}  \\ \hline
\end{tabular}
\caption{Number of variants included in the analysis per category}
\label{totTable}
\end{table}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
\caption{Frequencies of $Q$-associated SNPs shared between 1000G and either HapMap or GnomAD. The blue dot in the HapMap plot is the is the variant reported in \citep{Mandage2017}, as discussed in the text. There are 42 such SNPs in HapMap and 162 in the GnomAD.}
\label{HapMap_GnomAD}
\end{figure}

@@

%not necessarily all fake SNPs, there;s this bias, in most cases we expect them not to be real, there's a small chance that they are real
%could be false posittive, or have a strong calling bias
%majority of them don't reproduce, but a small portion of them do reproduce and should be looked at with care

%if the SNP was imputed with HapMap maybe it is real, but if this SNP is imputed using the other database, it might throw things off.

>>>>

Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques. 
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
For other analyses, removing suspicious SNPs may be adequate. 
Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.

@@

\section{Discussion}
Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data.
It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
However, considering that there are only \todo{number} suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}.
Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real, however we cannot discount the chance that they are real.
It is clear that the majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.

%\todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
%While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
%The data used was imputed genotype data using Hapmap II as the reference.
%One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ .
%While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
%Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
%We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.

@@

For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis. The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in a total of 80,971,616 variants ($S$) included in the test (see Table \ref{totTable}). We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.

\begin{table}[h]
\begin{tabular}{l  r r}
                      & {Repeat}  & {Non-Repeat}       \\ \hline
{SNP}  & 22,651,762 & 51,643,590  \\  
{INDEL} &  2,176,825  & 4,499,439 \\ \hline
\end{tabular}
\caption{Number of variants included in the analysis by category}
\label{totTable}
\end{table}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
\caption{Frequencies of $Q$-associated SNPs shared between 1000G and either HapMap or GnomAD. The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively. There are 70 such SNPs in HapMap and 186 in the GnomAD.}
\label{HapMap_GnomAD}
\end{figure}

@@
